---
title: "Day 2"
output:
  learnr::tutorial:
    progressive: true
    allow_skip: true
runtime: shiny_prerendered
editor_options:
  chunk_output_type: console
---

```{r setup, include=FALSE}

library(learnr)

tutorial_options(
  exercise.timelimit = 60,
  # A simple checker function that just returns the message in the check chunk
  exercise.checker = function(check_code, ...) {
    list(
      message = eval(parse(text = check_code)),
      correct = logical(0),
      type = "info",
      location = "append"
    )
  }
)
knitr::opts_chunk$set(error = TRUE)

r <- getOption("repos")
r["CRAN"] <- "https://cloud.r-project.org/"
options(repos = r)
```

## Let's recap Day 1

-   What are variables? Where can you see all the current variables in your workspace?

-   What are functions? Write out an example function.

```{r, recap0, exercise=TRUE, exercise.eval = FALSE}
```

-   How would you set the working directory in RStudio? Set the working directory for today's session.

-   What is the difference between a Vector and a DataFrame?

-   Get the second element of the vector `x` below

```{r, recap1, exercise=TRUE, exercise.eval = FALSE}
x <- c(3,6,8,0,-1,4)
```

-   Get all elements of `x` greater than 2

```{r, recap2, exercise=TRUE, exercise.eval = FALSE}
```

-   Find the sum of all elements of `x`

```{r, recap3, exercise=TRUE, exercise.eval = FALSE}
```

-   Where can you look up the documentation for R functions? Find help on the function `write.csv`

-   Install the `palmerpenguins` package and load it into the R workspace.

```{r, recap4, exercise=TRUE, exercise.eval = FALSE}
```

-   Find the names of all the columns in the `penguins` dataframe

```{r, recap5, exercise=TRUE, exercise.eval = FALSE}
```

-   Read in the file `ev_police_jan.csv` from the data folder

```{r, recap6, exercise=TRUE, exercise.eval = FALSE}
```

-   Install the `palmerpenguins` package and load it into the R workspace.

```{r, recap7, exercise=TRUE, exercise.eval = FALSE}
```
